---
title: "ClassifyR Package Vignette"
author: "Yibing Zhang, Zijin Li, Xinyi Hu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ClassifyR Package Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,      
  fig.height = 5,     
  out.width = "80%",  
  dpi = 300  
)
```

# 1. Package Overview

Our package is designed for binary classification problems. To demonstrate the complete workflow, we employ **four different datasets** and organize our **12 functions** into three categories:

1.  Data Preprocessing and Visualization
2.  Model Training
3.  Model Evaluation and Result Analysis

Target users

-   Students learning about machine learning classification algorithms
-   Researchers comparing multiple classification approaches on some dataset

# 2. Package Installation

To install this package, please use the command below and then load the package

```{r install-pkg, eval = FALSE}
install.packages("path/to/YourPackage_1.0.0.tar.gz", repos = NULL, type = "source")
```

```{r setup}
library(ClassifyR)
```

# 3. Data Preprocessing

Here we load our prepared datasets like this. We prepare 4 different datasets for your reference, and you can use your collected datasets.

```{r}
data("breast_cancer_dataset")
data("diabetes_dataset")
data("heart_dataset")
data("mushroom_dataset")
head(breast_cancer_dataset)
head(diabetes_dataset)
head(heart_dataset)
head(mushroom_dataset)
```

Then we use the function in preprocess.R to handle missing value, conduct correlation analysis, split train-test sets, conduct Min-Max normalization.

```{r}
heart_data_res = pre_process_data(heart_dataset, 14)
diabetes_data_res = pre_process_data(diabetes_dataset, 9)
mushroom_data_res = pre_process_data(mushroom_dataset, 1)

breast_cancer_data_res = preprocessBreastCancerData(breast_cancer_dataset)
```

```{r}
mushroom_data_exploratory_res = exploratory_data_analysis(mushroom_data_res$train_data, target_name ="class")

diabetes_data_exploratory_res = exploratory_data_analysis(diabetes_data_res$train_data,target_name = "Outcome")

heart_data_exploratory_res = exploratory_data_analysis(heart_data_res$train_data,target_name = "target")

breast_data_exploratory_res = exploratory_data_analysis(breast_cancer_data_res$train_data,target_name = "Status")
```

We can further use visualization helper function to better understand our datasets.

```{r}
# Breast Cancer Dataset
processed_data_visual(breast_cancer_data_res$train_data, target_col = "Status")

# Diabetes Dataser
processed_data_visual(diabetes_data_res$train_data, target_col = "Outcome")
```

Then after our statistical analysis and some visualization, we are getting to know our data and now we are ready to use them to train our model. In the next part we make a table to guide you to choose the proper model that suits your valuable data set.

# 4. Model Suitability for Different Training Sets

The following table provides a brief guideline on the characteristics of training sets that are generally well-suited for each of the five machine learning models implemented in the `ClassifyR` package.

| Model (Algorithm) | Ideal Training Set Characteristics | Considerations & Cautions |
|------------------|--------------------------|----------------------------|
| **SVM** (Support Vector Machine) | \- **Small to medium** number of samples (e.g., \< 10,000). <br> - **High-dimensional** data (many features, e.g., \> number of samples). <br> - Clear **margin of separation** between classes. <br> - Data with complex boundaries (using kernel trick). | \- Not ideal for very large datasets (slow training). <br> - Requires feature scaling/normalization. <br> - Sensitive to noisy data and outliers. <br> - Kernel selection (e.g., linear, RBF) is important. |
| **Decision Tree** | \- **Small to medium** datasets. <br> - Data with **mixed data types** (numeric, categorical). <br> - When **interpretability** (simple rules) is a priority. <br> - Features have non-linear and interactive effects. | \- Prone to **overfitting** on training data. <br> - **Unstable**; small changes in data can lead to a completely different tree. <br> - May not perform as well as ensemble methods on complex tasks. |
| **Random Forest** | \- **Medium to large** datasets. <br> - Data with **high dimensionality** (many features). <br> - Complex relationships between features. <br> - Datasets with **mixed data types**. <br> - Tolerant to outliers and noisy data. | \- Less **interpretable** than a single decision tree. <br> - Slower to train and predict than a single tree, but efficient for its accuracy. <br> - Memory-intensive for very large or high-dimensional data. |
| **XGBoost** | \- **Medium to very large** structured/tabular datasets. <br> - Data with complex, non-linear patterns. <br> - When **state-of-the-art predictive performance** is the goal. <br> - Datasets where feature interactions matter. | \- Requires careful **hyperparameter tuning** (learning rate, tree depth, subsample). <br> - Can **overfit** if not tuned properly. <br> - More sensitive to outliers than Random Forest. <br> - Slower to train than Random Forest for some datasets. |
| **Neural Network** | \- **Large datasets** (e.g., \> 10,000 samples). <br> - **High-dimensional** data (e.g., images, text, genomics). <br> - Extremely complex, non-linear underlying patterns. <br> - **Unstructured data** (e.g., images) when combined with specialized architectures (e.g., CNNs), but for tabular data, other models are often faster. | \- Requires **significant computational resources** (CPU/GPU). <br> - **Black-box** model; very hard to interpret. <br> - Requires careful architecture design and hyperparameter tuning. <br> - Often requires **feature scaling** and may benefit from dimensionality reduction. |

For the 4 data sets we prepared in this package, we use Neural Network for breast cancer data set. The parameters in this trainModel function should adjust your data set, you need to choose appropriate hidden neuron numbers and lr.

# 5. Model Training

## 5-1. Breast Cancer DataSet

For this dataset, we choose the neural network model.

```{r}
bc_model = trainModel(X=breast_cancer_data_res$X_train,y=breast_cancer_data_res$y_train,num_iteration = 5000,hidden_neurons = 15,lr=0.01,verbose = TRUE) # using verbose for training detail, you can turn it off by setting it to FALSE

plot(bc_model$cost_hist, type = "l", 
     xlab = "Iteration", ylab = "Cost",
     main = "Training Loss",
     col="red")

# we further use this model prediction result to evaluate this model
bc_NN_prediction_res= NN_performance(test_data = breast_cancer_data_res$X_test,hidden_neurons = 15,model_name = "Neural Network",train_model = bc_model)

```

## 5-2. Mushroom Dataset

For Mushroom data set, we use 2 models: random forests and decision tree.

```{r}
mushroom_rf_model <- train_rf(
  X = mushroom_data_res$X_train_norm,
  y = mushroom_data_res$y_train,
  n_trees = 20,
  max_depth = 5
)
mushroom_rf_result <- predict_rf_model(mushroom_rf_model, mushroom_data_res$X_test_norm)

mushroom_dt_model = decision_tree(mushroom_data_res$X_train_norm, mushroom_data_res$y_train, mushroom_data_res$X_test_norm, mushroom_data_res$y_test, cp = 0.01)
```

## 5-3. Heart Disease dataset

For Heart disease data set, we use 5 models, they are neural network, decision tree, XGBoost, SVM and random forests.

```{r}
# Neural Network
heart_NN_model = trainModel(
  X = t(as.matrix(heart_data_res$X_train_norm)),
  y = t(as.matrix(heart_data_res$y_train)),
  num_iteration = 7000,
  hidden_neurons = 10,
  lr = 0.3,
  verbose = FALSE
  ) 
plot(heart_NN_model$cost_hist, type = "l", 
     xlab = "Iteration", ylab = "Cost",
     main = "Training Loss",
     col="red")
heart_NN_result= NN_performance(
  test_data = t(as.matrix(heart_data_res$X_test)),
  hidden_neurons = 10,
  model_name = "Neural Network",
  train_model = heart_NN_model
  )

# Decision Tree
heart_dtr_result <- decision_tree(
  X_train = heart_data_res$X_train_norm,
  y_train = heart_data_res$y_train,
  X_test  = heart_data_res$X_test_norm,
  y_test  = heart_data_res$y_test
)

# XGBoost
heart_xgb_result <- train_xgb(
  X_train = heart_data_res$X_train_norm,
  y_train = heart_data_res$y_train,
  X_test  = heart_data_res$X_test_norm,
  y_test  = heart_data_res$y_test
)

# SVM
heart_svm_result <- train_svm(
  X_train = heart_data_res$X_train,
  y_train = heart_data_res$y_train,
  X_test  = heart_data_res$X_test,
  y_test  = heart_data_res$y_test,
  tune = TRUE
)

# Random Forest
heart_rf_model <- train_rf(
  X = heart_data_res$X_train_norm,
  y = heart_data_res$y_train,
  n_trees = 20,
  max_depth = 5
)
heart_rf_result <- predict_rf_model(heart_rf_model, heart_data_res$X_test_norm)
```

## 5-4. Diabetes dataset

For Diabetes data set, we use 5 models, they are neural network, decision tree, XGBoost, SVM and random forests.

```{r}
# Neural Network
diabetes_NN_model = trainModel(
  X = t(as.matrix(diabetes_data_res$X_train_norm)),
  y = t(as.matrix(diabetes_data_res$y_train)),
  num_iteration = 10000,
  hidden_neurons = 6,
  lr=0.6,
  verbose = FALSE
  ) 
plot(diabetes_NN_model$cost_hist, type = "l", 
     xlab = "Iteration", ylab = "Cost",
     main = "Training Loss",
     col="red")
diabetes_NN_result = NN_performance(
  test_data = t(as.matrix(diabetes_data_res$X_test)),
  hidden_neurons = 6,
  model_name = "Neural Network",
  train_model = diabetes_NN_model
  )

# Decision Tree
diabetes_dtr_result <- decision_tree(
  X_train = diabetes_data_res$X_train_norm,
  y_train = diabetes_data_res$y_train,
  X_test  = diabetes_data_res$X_test_norm,
  y_test  = diabetes_data_res$y_test
)

# XGBoost
diabetes_xgb_result <- train_xgb(
  X_train = diabetes_data_res$X_train_norm,
  y_train = diabetes_data_res$y_train,
  X_test  = diabetes_data_res$X_test_norm,
  y_test  = diabetes_data_res$y_test
)

# SVM
diabetes_svm_result <- train_svm(
  X_train = diabetes_data_res$X_train,
  y_train = diabetes_data_res$y_train,
  X_test  = diabetes_data_res$X_test,
  y_test  = diabetes_data_res$y_test,
  tune = TRUE
)

# Random Forest
diabetes_rf_model <- train_rf(
  X = diabetes_data_res$X_train_norm,
  y = diabetes_data_res$y_train,
  n_trees = 20,
  max_depth = 5
)
diabetes_rf_result <- predict_rf_model(diabetes_rf_model, diabetes_data_res$X_test_norm)
diabetes_rf_predictions <- diabetes_rf_result$predictions
diabetes_rf_pred_prob <- diabetes_rf_result$pred_prob
```

# 6. Model Evaluation and Result Analysis

## 6-1. Breast Cancer dataset

```{r}
evaluate_models(breast_cancer_data_res$y_test, bc_NN_prediction_res,"1")
```

**Performance Analysis:** The single-model evaluation function `evaluate_models` provides a comprehensive view of the Neural Network's performance. The overall Accuracy and F1-Score are robust. The model achieved a high Area Under the Curve (AUC) and the heatmap (confusion matrix) reveals a large proportion of True Positives and True Negatives, indicating a strong ability to distinguish between classes.

## 6-2. Mushroom dataset

```{r}
evaluate_models(mushroom_data_res$y_test, mushroom_rf_result,"1")
evaluate_models(mushroom_data_res$y_test, mushroom_dt_model, "1")
```

Since we use 2 different models for the mushroom prediction, we can use the function below to compare these 2 models

```{r}
mushroom_models_com = list(
  "Decision Tree" = list(predictions = mushroom_dt_model$predictions,
                         pred_prob=mushroom_dt_model$pred_prob),
  "Random Forest" = list(predictions=mushroom_rf_result$predictions,
                         pred_prob = mushroom_rf_result$pred_prob))

models_comparison(mushroom_data_res$y_test, mushroom_models_com, "1")
```

**Performance Analysis and Comparison:** Because the Mushroom dataset relies heavily on categorical logic rules (e.g., odor, cap shape), both models perform well. While the standalone Decision Tree provides excellent interpretability, the Random Forest model mitigates the risk of overfitting by aggregating multiple trees. Consequently, the Random Forest output demonstrates higher stability and a near-perfect F1-Score, proving it to be the superior algorithm for this specific data structure.

## 6-3. Heart Disease dataset

```{r}
heart_models_com = list(
  "Neural Network" = heart_NN_result,
  "Decision Tree" = heart_dtr_result,
  "XGBoost" = heart_xgb_result,
  "Random Forest" = heart_rf_result,
  "SVM Result" = heart_svm_result
  )
models_comparison(heart_data_res$y_test, heart_models_com, "1")
```

**Performance Analysis and Comparison:** Random forest strikes the best balance between classes in the confusion matrix and achieves the highest overall accuracy and AUC, demonstrating strong class separation. XGBoost excels at identifying actual positive cases, which is highly desirable in medical contexts to avoid missing sick patients. However, this high sensitivity comes at the expense of lower precision (more false positives). SVM delivers highly balanced and competitive results. It achieves the highest precision, meaning its positive predictions are the most reliable among all models. Decision Tree performs moderately well across all metrics. While its overall performance is slightly edged out by ensemble methods, it provides a transparent and easily interpretable baseline. Neural Network struggles on this dataset, lagging behind in all major metrics. This indicates that the current architecture or hyperparameters require further tuning to effectively capture the non-linear patterns here.

## 6-4. Diabetes Dataset

```{r}
diabetes_models_com = list(
  "Neural Network" = diabetes_NN_result,
  "Decision Tree" = diabetes_dtr_result,
  "XGBoost" = diabetes_xgb_result,
  "Random Forest" = diabetes_rf_result,
  "SVM Result" = diabetes_svm_result
  )
models_comparison(diabetes_data_res$y_test, diabetes_models_com, "1")
```

**Performance Analysis and Comparison:** Overall, SVM and random Forest are best. SVM provides the most balanced performance across all core metrics. Random Forest minimizes false positives and making it highly reliable when predicting a positive diabetes case. XGBoost identifies a large majority of actual positive cases. If the medical priority is to minimize missed diagnoses, XGBoost is an excellent choice, despite generating more false positives. Decision Tree performs well and proves that simple, interpretable rule-based splits are quite effective for the clinical thresholds inherent in this noisy dataset. Neural Network fails on this dataset. Due to overly aggressive tuning, the model failed to converge and acts almost as a constant predictor. This inflates recall but completely destroys precision.

# 7. The advantage of our package

1.  It demonstrates the complete pipeline for binary classification, including data preparation, data cleaning, exploratory data analysis, model training, prediction making, model comparison, and data/model visualization.

2.  It includes five commonly used machine learning classification methods that are suitable for a wide range of datasets.

3.  Well-organized helper functions provide a clear and accessible guide for machine learning and R language learners on how to use our package effectively.

# 8. The points learnt in this course

1.  R Function Development: How to design, implement, and structure R functions with proper parameters, error handling, and return values.

2.  R Package Ecosystem: How to leverage powerful packages including ggplot2 (visualization), caret (machine learning workflows), and GGally (exploratory analysis).

3.  Function Coordination: How to coordinate multiple functions and design clean internal interfaces to build a cohesive and maintainable R package.

4.  Package Structure & Development: How to structure an R package (DESCRIPTION, NAMESPACE, R/ folder, data/ folder) and use devtools for efficient development.

5.  Documentation & Vignettes: How to document functions properly using roxygen2 and build complete package vignettes to help users understand and adopt the package.

6.  Data Processing & Exploration: How to preprocess real-world datasets (handling missing values, scaling, data splitting) and conduct exploratory data analysis.

7.  Model Evaluation & Comparison: How to evaluate classification models using standardized metrics (accuracy, precision, recall, F1, AUC-ROC) and compare multiple models.

8.  Data & Model Visualization: How to create publication-quality visualizations for exploratory analysis, model comparison, and performance reporting.

9.  Version Control & Collaboration: How to use Git and GitHub for version control and collaborative package development.

10. Testing & Quality Assurance: How to write unit tests using testthat and perform package checks to ensure code reliability and robustness.

# 9. Conclusion

The ClassifyR package provides a comprehensive and user-friendly framework for binary classification tasks, integrating essential steps from data preprocessing to model evaluation and comparison. The package implements five widely-used machine learning algorithms—Support Vector Machine (SVM), Decision Tree, Random Forest, XGBoost, and Neural Network—allowing users to apply and compare multiple approaches on their datasets with ease.

Key features of the package include:

-   Streamlined data preprocessing: Functions for handling missing values, correlation analysis, train-test splitting, and Min-Max normalization ensure that datasets are properly prepared before model training.

-   Exploratory data analysis and visualization: Built-in functions enable users to understand data distributions and feature relationships through statistical summaries and publication-quality visualizations.

-   Multiple modeling approaches: Five distinct classification algorithms are provided, each with a consistent interface, making it straightforward to train models and generate predictions.

-   Comprehensive model evaluation: The package offers standardized performance metrics including accuracy, precision, recall, F1-score, and AUC-ROC, along with ROC curve visualization.

-   Model comparison capabilities: Users can easily compare the performance of multiple models on the same dataset using the dedicated models_comparison() function.

-   Modular design: All data processing, model training, prediction, and evaluation steps are organized into independent, well-documented functions, ensuring code clarity, reusability, and extensibility.

Through the four example datasets (Breast Cancer, Diabetes, Heart Disease, and Mushroom), users can follow a complete classification workflow and adapt it to their own data. The package also incorporates unit testing to ensure reliability and robustness.

In summary, ClassifyR serves as both a practical tool for data analysts and machine learning engineers and an educational resource for students learning classification techniques in R. Its consistent interface and modular structure lower the barrier to entry for machine learning workflows while maintaining the flexibility needed for real-world applications.

For more information, please refer to the function documentation or explore the examples in this vignette.
