---
title: "VignetteName"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VignetteName}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ClassifyR)
```

# Package Overview

Our package is designed for binary classification problems. To demonstrate the complete workflow, we employ **four different datasets** and organize our **12 functions** into three categories:

1.  Data Preprocessing and Visualization
2.  Model Training
3.  Model Evaluation and Result Analysis

## Data Preprocessing
Here we load our prepared data sets and load them like this.
```{r}
data("breast_cancer_dataset")
data("diabetes_dataset")
data("heart_dataset")
data("mushroom_dataset")
head(breast_cancer_dataset)
head(diabetes_dataset)
head(heart_dataset)
head(mushroom_dataset)
```
Then we use the function in preprocess.R to handle missing value, do correlation analysis, split train-test sets, do Min-Max normalization. 
```{r}
heart_data_res = pre_process_data(heart_dataset, 14)
diabetes_data_res = pre_process_data(diabetes_dataset,8)
mushroom_data_res = pre_process_data(mushroom_dataset,1)

breast_cancer_data_res = preprocessBreastCancerData(breast_cancer_dataset)
```

